Classical Simulation of Quantum Error Correction in a Fibonacci Anyon Code

### 2015-05-06 fib2.tex
edit on skype with steve. addressed a bunch of comments, added some refs, other minor stuff.

We use a phenomenological model of Fibonacci anyon dynamics which neglects any 
microscopic details of the system. 
\dude{Another way of saying this is that we are doing a tqft simulation.}
This is consistent with the principles of topologically 
ordered systems and anyonic physics, where the key universal features describing the 
anyon model correspond to large length-scale physics, while the microscopic physics plays 
a less important (and non-universal) role. As such, we model the system simply as a 
periodic $L\times L$ square lattice $\Lambda$ of sites upon which sets of anyons can 
reside. Allowed anyon dynamics (pair-creation, braiding, and fusion) are considered to act 
locally on edges of $\Lambda$. For more details of an analogous model, see~\cite{Brell2013}.

@@

In order to perform a logical error on our code, a noise process must have support on a 
ribbon running around a homologically nontrivial loop of the lattice. These correspond to 
processes in which anyonic charge is transported around the non-trivial loop before 
annihilating to vacuum \dude{or not}.

Our error-correction algorithm is based on a hard-decision renormalization group (RG) 
decoder~\cite{Bravyi2011}. 
The decoder proceeds by measuring the occupations of 
anyons at each site of the lattice, forming clusters of nearby anyons and fusing the anyons 
within each cluster, before iterating at increasing length scales.
It terminates when the lattice 
is free of anyons (i.e.~we have returned to the codespace) or the clustering length scale 
exceeds the total size of the lattice.
\cggb{A few more details? What is our length schedule?} 
\cggb{How does our decoder relate to other decoding schemes used in~\cite{Wootton2013,Brell2013,Hutter2014}?} 
Note that, as was found in Ref.~\cite{Brell2013}, we expect that our qualitative results may be reproduced by most alternative families of decoders, such as matching decoders~\cite{?} and soft-decisions 
RG decoders~\cite{?} \cggb{do we?}.
The advantage of using a hard-decision RG decoder is 
its simplicity and flexibility, and the fact that its clustering scheme is compatible with the structure in the noise processes that allows us to classically simulate them.

\begin{figure*}[th!]
\begin{center}
	{\scriptsize (i)
	\includegraphics[width=0.57\columnwidth]{pair-create.pdf}
	\hskip 4pt
	(ii)
	\includegraphics[width=0.57\columnwidth]{syndrome-1.pdf}
	\hskip 4pt
	(iii)}
	\includegraphics[width=0.57\columnwidth]{syndrome-2.pdf}
\caption{The simulation of noise processes proceeds by i) creating pairs of anyons on 
neighbouring tiles represented by open curves ii) joining the curves within a given tile 
iii) measuring the total charge on each tile.}
\label{f:simex}
\end{center}
\end{figure*}

@@

Below the (bond) percolation threshold for (say) a 2D square lattice, we expect random sets of 
bonds to decompose into separate clusters of average size $O(\log(n))$ and variance 
$O(1)$~\cite{Bazant2000}. \cggb{Do we want to use the word cluster here again or switch? What was our alternative word again?} 
Each noise process in our model is associated with a (randomly distributed) edge, and so 
disconnected clusters correspond to sets of anyons that could not have interacted at any 
point in their history. As such, we can simulate the braiding processes of each cluster 
separately. In other words: the quantum state in the fusion space of all anyons factorizes into 
a tensor product over bond clusters, and we can compute each factor separately. Since each 
cluster has size only $O(\log(n))$, we can typically simulate these dynamics efficiently 
because the resulting fusion space has dimension $O(\mathrm{poly}(n))$. There are rare 
random processes that violate this reasoning, but processes such as these are suppressed 
exponentially in $L$, the lattice size~\cite{Grimmett1989}.

@@

Of course, if we were to consider noise so strong that the clusters percolate over the lattice, 
we would no longer be able to efficiently simulate the system. However, recall that logical 
errors in our system correspond to processes that act over a non-trivial loop of the torus. If 
the noise operations are so common to have percolated, this is an example of a situation 
where we expect error-correction \emph{not} to succeed. Similarly, if the noise has not yet 
percolated by itself, but the clusters are merged into a large (percolated) cluster by the 
decoding algorithm, this is again an example of a process that we expect to correspond to a 
failure of error-correction. In this way, we see that the events that we are able to efficiently 
simulate correspond precisely to those for which we expect error-correction to succeed. We 
can simply declare failure on non-simulable events. This allows us to simulate and 
demonstrate the success of error-correction schemes for Fibonacci anyons, without explicitly 
simulating the failure of such schemes.

@@

Our concrete simulation proceeds by associating a \emph{tile} with each site of the 
$L\times L$ periodic square lattice.
In order to keep track of the observables of the system 
we maintain a set of disjoint curves, representing the history of anyons on the lattice.
These curves can also be thought of as a top-down view of the fusion tree:
the anyons on the leaves of the fusion tree are here displayed as points along each curve.
%The ends of curves correspond to leaves of the fusion tree, or anyons. 
In general, the curves wind around the two 
dimensional tiles in a haphazard way determined by the progress of the simulation.
This allows a dynamically generated basis for the fusion space that easily allows for tensor 
decomposition into factors corresponding to disjoint clusters.
For comparison, previous work 
used a single fixed basis for the fusion space (determined by row-major ordering of 
sites)~\cite{Brell2013}.

@@

\begin{enumerate}[i)]
\item The noise process is modelled by a sequence of pair-creation events distributed as 
a Poisson process across each edge. Curves are created crossing the relevant edge 
and terminating on the neighbouring sites.
\item We first join curves that participate in the same tile in order to, 
\cggb{maybe a couple more lines of detail on this and the next point?}
\item measure the charge on each tile. Any resulting anyon is left somewhere in the tile.
\item The decoder examines this syndrome and clusters nearby anyons. Each cluster 
is then measured (fused), and the decoder continues to cluster and fuse on larger 
scales until there are no more anyons, or a cluster includes a non-trivial loop on the 
lattice. Since operators with support on such non-trivial clusters will generally perform an 
undetectable error on the encoded state, in this case we abandon the simulation and 
declare failure to error correct.
\end{enumerate}

@@

We plot the performance of the decoder as a function of error rate for varying lattice sizes in 
\Fref{f:threshold}. The error rate is parameterized by the Poisson process duration 
$t_{\mathrm{sim}}$, representing the expected number of errors per edge over the simulation. 
We find evidence of a decoding threshold below which decoding succeeds with asymptotic 
certainty as the system size increases at $t_{\mathrm{sim}}\simeq 0.125 \pm 0.003$.

@@

\cggb{Do we need to define $t_{\mathrm{sim}}$ more explicitly?}

\cggb{Comment on the effect of declaring failure on some simulations on this plot. Would 
we expect it to change if we simulated every event completely? If so, how?}
\dude{you spend alot of effort arguing above that we can
neglect these occurances, but i am wondering if we need to be
this strenuous: i did many calculations of global braids/fusions
and they all do a non-unitary on the logical space.. i could not
see how to braid around the torus in *any* trivial way *at all*.
This falls short of a proof, but i'm fairly convinced our sim is
in fact 100 percent correct.}

@@

We have demonstrated classical simulation of successful error correction in a universal anyon 
model. Though we have chosen several properties of our model and simulation in a 
convenient way for simplicity, we think it very unlikely that they will affect the qualitative 
results, similarly to the results of Ref.~\cite{Brell2013}. In particular, although we have 
modelled our logical qubit as encoded in the global topological degrees of freedom of our 
system, we could have encoded it in the fusion space of several preferred $\tau$ anyons. 
This situation would be appropriate to model error-correction routines in a topological 
quantum computation protocol. Additionally, we expect our results to be stable to changes 
in details of the noise model, decoding algorithm, etc.

None of our techniques are restricted to simulation of Fibonacci anyon dynamics, and could 
equally well be used to simulate successful error-correction protocols in an arbitrary anyon 
code.

@@

\stf{Is there a possibility of using MPS like Rob Pfeiffer to improve the simulations or 
include interactions?}~\cite{Pfeifer2010, Pfeifer2012}

>>>>

We use a phenomenological model of Fibonacci anyon dynamics which neglects any 
microscopic details of the system. 
This is consistent with the principles of topologically 
ordered systems and anyonic physics, where the key universal features describing the 
anyon model correspond to large length-scale physics, while the microscopic physics plays 
a less important (and non-universal) role. Effectively, we simulate a topological quantum field theory on a discrete space-time.
As such, we model the system simply as a 
periodic $L\times L$ square lattice $\Lambda$ of sites upon which sets of anyons can 
reside. Allowed anyon dynamics (pair-creation, braiding, and fusion) are considered to act 
locally on edges of $\Lambda$. For more details of an analogous model, see~\cite{Brell2013}.

@@

In order to perform a logical error on our code, a noise process must have support on a 
ribbon running around a homologically nontrivial loop of the lattice. These correspond to 
processes in which anyonic charge is transported around the non-trivial loop before 
annihilating to vacuum.
% cggb{we want to define logical operators as those that map the codespace to itself, so for it to be a logical operator, they must annihilate to vacuum. they need not be unitary}.

Our error-correction algorithm is based on a hard-decision renormalization group (RG) 
decoder~\cite{Bravyi2011}. 
The decoder proceeds by measuring the occupations of 
anyons at each site of the lattice, forming clusters of nearby anyons and fusing the anyons 
within each cluster, before iterating at increasing length scales.
It terminates when the lattice 
is free of anyons (i.e.~we have returned to the codespace) or the clustering length scale 
exceeds the total size of the lattice.
\cggb{A few more details? What is our length schedule?} 
\cggb{How does our decoder relate to other decoding schemes used in~\cite{Wootton2013,Brell2013,Hutter2014}?} 
Note that, as was found in Ref.~\cite{Brell2013}, we expect that our qualitative results may be reproduced by most alternative families of decoders, although it may not be clear in all cases how to extend these decoders to the non-abelian setting.
The advantage of using a hard-decision RG decoder is 
its simplicity and flexibility, and the fact that its clustering scheme is compatible with the structure in the noise processes that allows us to classically simulate them.

\begin{figure*}[th!]
\begin{center}
	{\scriptsize (a)
	\includegraphics[width=0.57\columnwidth]{pair-create.pdf}
	\hskip 4pt
	(b)
	\includegraphics[width=0.57\columnwidth]{syndrome-1.pdf}
	\hskip 4pt
	(c)}
	\includegraphics[width=0.57\columnwidth]{syndrome-2.pdf}
\caption{The simulation of noise processes proceeds by a) creating pairs of anyons on 
neighbouring tiles represented by open curves b) joining the curves within a given tile 
c) measuring the total charge on each tile.}
\label{f:simex}
\end{center}
\end{figure*}

@@

Below the (bond) percolation threshold for (say) a 2D square lattice, we expect random sets of 
bonds to decompose into separate clusters~\footnote{Note that we use the word cluster to refer to both groupings of nearby anyons as formed by the decoding algorithm, and also to connected components of a graph under percolation, since both of these usages are standard in the literature. This distinction should be kept in mind, though context should remove any ambiguity regarding the meaning.} of average size $O(\log(n))$ and variance 
$O(1)$~\cite{Bazant2000}.
Each noise process in our model is associated with a (randomly distributed) edge, and so 
disconnected clusters correspond to sets of anyons that could not have interacted at any 
point in their history. As such, we can simulate the braiding processes of each cluster 
separately. In other words: the quantum state in the fusion space of all anyons factorizes into 
a tensor product over bond clusters, and we can compute each factor separately. Since each 
cluster has size only $O(\log(n))$, we can typically simulate these dynamics efficiently 
because the resulting fusion space has dimension $O(\mathrm{poly}(n))$. There are rare 
random processes that violate this reasoning, but processes such as these are suppressed 
exponentially in $L$, the lattice size~\cite{Grimmett1989}.

@@

Of course, if we were to consider noise so strong that the clusters percolate over the lattice, 
we would no longer expect to be able to efficiently simulate the system. However, recall that logical 
errors in our system correspond to processes that act over a non-trivial loop of the torus. If 
the noise operations are so common to have percolated, this is an example of a situation 
where we expect error-correction \emph{not} to succeed. Similarly, if the noise has not yet 
percolated by itself, but the clusters are merged into a large (percolated) cluster by the 
decoding algorithm, this is again an example of a process that we expect in general to correspond to a 
failure of error-correction. In this way, we see that the events that we are able to efficiently 
simulate correspond to those for which we expect error-correction to succeed. We 
can simply declare failure on non-simulable events. This allows us to simulate and 
demonstrate the success of error-correction schemes for Fibonacci anyons, without explicitly 
simulating the failure of such schemes.

@@

Our concrete simulation proceeds by associating a \emph{tile} with each site of the 
$L\times L$ periodic square lattice.
In order to keep track of the observables of the system 
we maintain a set of disjoint curves, representing the history of anyons on the lattice.
These curves can also be thought of as a top-down view of the fusion tree:
the anyons comprising the fusion tree are here displayed as points along each curve. \cggb{Is this still true?}
%The ends of curves correspond to leaves of the fusion tree, or anyons. 
In general, the curves wind around the two 
dimensional tiles in a haphazard way determined by the progress of the simulation.
This allows a dynamically generated basis for the fusion space that easily allows for tensor 
decomposition into factors corresponding to disjoint clusters.
For comparison, previous work 
used a single fixed basis for the fusion space (determined by row-major ordering of 
sites)~\cite{Brell2013}.

@@

\begin{enumerate}[i)]
\item The noise process is modelled by a sequence of pair-creation events distributed as 
a Poisson process across each edge. Curves are created crossing the relevant edge 
and terminating on the neighbouring sites.
\item We first join curves that participate in the same tile in order to, 
\cggb{maybe a couple more lines of detail on this and the next point? perhaps also the previous paragraph?}
\item measure the charge on each tile. Any resulting anyon is left somewhere in the tile.
\item The decoder examines this syndrome and clusters nearby anyons. Each cluster 
is then measured (fused), and the decoder continues to cluster and fuse on larger 
scales until there are no more anyons, or a cluster includes a non-trivial loop on the 
lattice. Since operators with support on such non-trivial clusters will generally perform an 
undetectable error on the encoded state, in this case we abandon the simulation and 
declare failure to error correct.
\end{enumerate}

@@

We plot the performance of the decoder as a function of error rate for varying lattice sizes in 
\Fref{f:threshold}. The error rate is parameterized by the Poisson process duration 
$t_{\mathrm{sim}}$, representing the expected number of errors per edge during the simulation. 
We find evidence of a decoding threshold below which decoding succeeds with asymptotic 
certainty as the system size increases at $t_{\mathrm{sim}}\simeq 0.125 \pm 0.003$.

@@

%\cggb{Do we need to define $t_{\mathrm{sim}}$ more explicitly?}

Although we can guarantee that error-correction will succeed whenever noise does not percolate, it is possible that percolated events may still not result in an error. The connection between the percolation threshold and the error-correction threshold is not well understood in general~\cite{Hastings2014}, though it is clear that our threshold estimate will be a lower bound for the true threshold that may have been found if all events (including those that have percolated) were simulated. However, by simulating such percolated events on small lattices, our evidence suggests that for the Fibonacci anyon code that we study, the vast majority of such events will result in a logical error. As such we feel confident that neglecting such events does not have a significant effect on the threshold value obtained.

@@

We have demonstrated classical simulation of successful error correction in a universal anyon 
model. Though we have chosen several properties of our model and simulation in a 
convenient way for simplicity, we think it very unlikely that they will affect the qualitative 
results, similarly to the results of Ref.~\cite{Brell2013}. In particular, although we have 
modelled our logical qubit as encoded in the global topological degrees of freedom of our 
system, we could have encoded it in the fusion space of several preferred $\tau$ anyons. 
This situation would be appropriate to model error-correction routines in a topological 
quantum computation protocol. Additionally, we expect our results to be stable to changes 
in details of the noise model, decoding algorithm, and so on.

None of our techniques are restricted to simulation of Fibonacci anyon dynamics, and could 
equally well be used to simulate successful error-correction protocols in an arbitrary anyon 
code. As such, our methods could be used to demonstrate successful error-correction in an arbitrary anyonic topological quantum computation.

Applying our methods to more realistic models such as concrete microscopic spin models or models with non-topological features is an open question.

@@

\stf{Is there a possibility of using MPS like Rob Pfeifer to improve the simulations or 
include interactions?}~\cite{Pfeifer2010, Pfeifer2012, Singh2014}
